---
title: "Data Wrangling Project"
author: "Jyoti Chaudhary"
date: "December 6, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(rvest)
library(plyr)
library(stringr)
library(tidyr)
library(dplyr)
library(broom)
library(readxl)
library(ggplot2)
library(faraway)
library(choroplethrMaps)
library(choroplethr)
library(RColorBrewer)
```

Load excel data sheet "National_salary.xlsx" into R. This data sheet has US National salary data for all major occupations. The code extracts the salary details of major occupations and will plot a density graph with Annual Mean Salary plotted on X axis and Employment count for each occupation plotted on Y axis.


```{r message=FALSE, warning=FALSE, echo=TRUE}
nat_salary <- read_excel("National_salary.xlsx", na="", col_names = TRUE, col_types = NULL)

colnames(nat_salary) <- c("Code", "Job.title", "Group", "Emp.count", "Emp.PRSE", "Mean.hourly.wage", "Mean.annual.wage", "Mean.PRSE", "H.PCT10", "H.PCT25", "Median.hourly.wage", "H.PCT75", "H.PCT90", "A.PCT10", "A.PCT25", "Annual.median.wage", "A.PCT75", "A.PCT90", "Annual", "Hourly")

## Change numberic fields from 'character' class to 'numeric' class ##################

nat_salary$Mean.hourly.wage <- fround(as.numeric(nat_salary$Mean.hourly.wage), 2)
nat_salary$Median.hourly.wage <- fround(as.numeric(nat_salary$Median.hourly.wage), 2)
nat_salary$H.PCT10 <- fround(as.numeric(nat_salary$H.PCT10), 2)
nat_salary$H.PCT25 <- fround(as.numeric(nat_salary$H.PCT25), 2)
nat_salary$H.PCT75 <- fround(as.numeric(nat_salary$H.PCT75), 2)
nat_salary$H.PCT90 <- fround(as.numeric(nat_salary$H.PCT90), 2)
nat_salary$A.PCT10 <- fround(as.numeric(nat_salary$A.PCT10), 2)
nat_salary$A.PCT25 <- fround(as.numeric(nat_salary$A.PCT25), 2)
nat_salary$A.PCT75 <- fround(as.numeric(nat_salary$A.PCT75), 2)
nat_salary$A.PCT90 <- fround(as.numeric(nat_salary$A.PCT90), 2)
nat_salary <- nat_salary %>% mutate(code2 = substr(nat_salary$Code, 1, 2), major = "")

nat_salary$major <- ifelse(nat_salary$code2 == "11", c("Management"), nat_salary$major)
nat_salary$major <- ifelse(nat_salary$code2 == "13", c("Business and Financial Operations"), nat_salary$major)  
nat_salary$major <- ifelse(nat_salary$code2 == "15", c("Computer and Mathematical"), nat_salary$major)
nat_salary$major <- ifelse(nat_salary$code2 == "17", c("Architecture and Engineering Occupations"), nat_salary$major)  
nat_salary$major <- ifelse(nat_salary$code2 == "19", c("Life, Physical, and Social Science"), nat_salary$major)  
nat_salary$major <- ifelse(nat_salary$code2 == "21", c("Community and Social Service"), nat_salary$major)
nat_salary$major <- ifelse(nat_salary$code2 == "23", c("Legal"), nat_salary$major)

nat_salary$major <- ifelse(nat_salary$code2 == "25", c("Education, Training, and Library"), nat_salary$major)
nat_salary$major <- ifelse(nat_salary$code2 == "27", c("Arts, Design, Entertainment, Sports, and Media"), nat_salary$major)
nat_salary$major <- ifelse(nat_salary$code2 == "29", c("Healthcare Practitioners"), nat_salary$major)
nat_salary$major <- ifelse(nat_salary$code2 == "31", c("Healthcare Support"), nat_salary$major)
nat_salary$major <- ifelse(nat_salary$code2 == "33", c("Protective Service"), nat_salary$major)
nat_salary$major <- ifelse(nat_salary$code2 == "35", c("Food Preparation and Serving"), nat_salary$major)
nat_salary$major <- ifelse(nat_salary$code2 == "37", c("Building and Grounds Cleaning and Maintenance"), nat_salary$major)
nat_salary$major <- ifelse(nat_salary$code2 == "39", c("Personal Care and Service"), nat_salary$major)
nat_salary$major <- ifelse(nat_salary$code2 == "41", c("Sales and Related"), nat_salary$major)
nat_salary$major <- ifelse(nat_salary$code2 == "43", c("Office and Administrative Support"), nat_salary$major)
nat_salary$major <- ifelse(nat_salary$code2 == "45", c("Farming, Fishing, and Forestry"), nat_salary$major)
nat_salary$major <- ifelse(nat_salary$code2 == "47", c("Construction and Extraction"), nat_salary$major)
nat_salary$major <- ifelse(nat_salary$code2 == "49", c("Installation, Maintenance, and Repair"), nat_salary$major)
nat_salary$major <- ifelse(nat_salary$code2 == "51", c("Production"), nat_salary$major)
nat_salary$major <- ifelse(nat_salary$code2 == "53", c("Transportation and Material Moving"), nat_salary$major)


```


```{r message=FALSE, warning=FALSE}

salary1 <- select(nat_salary, Code, Job.title, Group, Emp.count, Mean.hourly.wage, Mean.annual.wage, Median.hourly.wage, Annual.median.wage)

salary_major <- salary1 %>% filter(Group=="major") 

salary_major_detail <- nat_salary %>% filter(Group=="major" | Group=="detailed") %>% select(Code, code2, Job.title, major, Group, Emp.count, Mean.hourly.wage, Mean.annual.wage, Median.hourly.wage, Annual.median.wage) 

```

```{r echo=TRUE, message=FALSE, warning=FALSE}
salary_major_detail %>% filter(Group != 'major') %>%
  ggplot(aes(x = Mean.annual.wage)) +
 geom_density(aes(fill = major), alpha = 0.6) +
  facet_wrap(~major, scales="free_y") + theme(legend.position="none")
```

SALARY OF TOP 10 PAYING OCCUPATIONS

```{r message=FALSE, warning=FALSE}
top10_salary1 <- nat_salary %>%
  filter(Group=="detailed") %>%
  arrange(desc(Mean.annual.wage))%>%
  head(10) %>%
  select(Job.title, Mean.annual.wage, A.PCT10, A.PCT25, Annual.median.wage, A.PCT75, A.PCT90) 

top10_salary <- select(top10_salary1, Job.title, Mean.annual.wage)

head(top10_salary)


```

SALARY OF LOWEST 10 PAYING OCCUPATIONS

```{r message=FALSE, warning=FALSE}
low10_salary1 <- nat_salary %>%
  filter(Group=="detailed" & Mean.annual.wage != "NA") %>%
  arrange(desc(Mean.annual.wage))%>%
  tail(10) %>%
  select(Job.title, Mean.annual.wage, A.PCT10, A.PCT25, Annual.median.wage, A.PCT75, A.PCT90)  

low10_salary <- select(low10_salary1, Job.title, Mean.annual.wage)

low10_salary$Job.title <- substr(low10_salary$Job.title, 1, 25)
  
head(low10_salary)

```


BOXPLOT FOR SALARY RANGE OF TOP PAYING OCCUPATIONS:  Anesthesiologists, Chief Executives and Family and General Practitioners

```{r message=FALSE, warning=FALSE}

top10_salary1 <- filter(top10_salary1, Job.title == "Anesthesiologists" | Job.title == "Family and General Practitioners" | Job.title =="Chief Executives")
  
top_low_salary <- rbind(top10_salary1, low10_salary1)

top_low_salary$A.PCT10 <- as.numeric(top_low_salary$A.PCT10)
top_low_salary$A.PCT25 <- as.numeric(top_low_salary$A.PCT25)
top_low_salary$A.PCT75 <- as.numeric(top_low_salary$A.PCT75)
top_low_salary$A.PCT90 <- as.numeric(top_low_salary$A.PCT90)

top_low_salary$Job.title <- substr(top_low_salary$Job.title, 1, 15)

```

```{r message=FALSE, warning=FALSE, echo=TRUE}

ggplot(top_low_salary[1:3,], aes(Job.title)) +
  geom_boxplot(fill="lightblue",
    aes(ymin = A.PCT10, lower = A.PCT25, middle = Annual.median.wage, upper = A.PCT75, ymax = A.PCT90),
    stat = "identity") + theme_light(base_size=16) + ylab("Annual Salary ($)") + xlab("Job Title")

```

BOXPLOT FOR SALARY RANGE OF LOWEST PAYING OCCUPATIONS

```{r message=FALSE, warning=FALSE, echo=TRUE}

ggplot(top_low_salary[4:7,], aes(Job.title)) +
  geom_boxplot(fill="lightblue",
    aes(ymin = A.PCT10, lower = A.PCT25, middle = Annual.median.wage, upper = A.PCT75, ymax = A.PCT90),
    stat = "identity"
  ) + theme_light(base_size=16) + ylab("Annual Salary ($)") + xlab("Job Title") 

```


Load excel data sheet "state_May2015_dl.xlsx" into R. This data sheet has US income wage data, by State, for all major occupations. The code extracts the salary details of major occupations and will plot a Horizontal Bar graph for total employment count by state. 

```{r message=FALSE, warning=FALSE}

state_salary <- read_excel("state_May2015_dl.xlsx", na="", col_names = TRUE, col_types = NULL)


state_salary1 <- select(state_salary, ST, STATE, OCC_CODE,  OCC_TITLE, OCC_GROUP, TOT_EMP, LOC_Q, H_MEAN, A_MEAN, H_MEDIAN, A_MEDIAN)

colnames(state_salary1) <- c("ST.code", "State", "OCC.Code",  "Job.title", "Group", "Emp.count", "LOC.Q", "H.mean.wage", "A.mean.wage", "H.median.wage", "A.median.wage")


state_salary1$LOC.Q <- fround(as.numeric(state_salary1$LOC.Q), 2)
state_salary1$H.mean.wage <- fround(as.numeric(state_salary1$H.mean.wage), 2)
state_salary1$A.mean.wage <- fround(as.numeric(state_salary1$A.mean.wage), 2)
state_salary1$H.median.wage <- fround(as.numeric(state_salary1$H.median.wage), 2)
state_salary1$A.median.wage <- fround(as.numeric(state_salary1$A.median.wage), 2)

state_salary1 <- state_salary1 %>% mutate(code2 = substr(state_salary1$OCC.Code, 1, 2), major = "")

state_salary_major <- state_salary1 %>% filter(Group=="major") 


state_salary1$major <- ifelse(state_salary1$code2 == "11", c("Management"), state_salary1$major)
state_salary1$major <- ifelse(state_salary1$code2 == "13", c("Business and Financial Operations"), state_salary1$major)  
state_salary1$major <- ifelse(state_salary1$code2 == "15", c("Computer and Mathematical"), state_salary1$major)
state_salary1$major <- ifelse(state_salary1$code2 == "17", c("Architecture and Engineering Occupations"), state_salary1$major)  
state_salary1$major <- ifelse(state_salary1$code2 == "19", c("Life, Physical, and Social Science"), state_salary1$major)  
state_salary1$major <- ifelse(state_salary1$code2 == "21", c("Community and Social Service"), state_salary1$major)
state_salary1$major <- ifelse(state_salary1$code2 == "23", c("Legal"), state_salary1$major)
state_salary1$major <- ifelse(state_salary1$code2 == "25", c("Education, Training, and Library"), state_salary1$major)
state_salary1$major <- ifelse(state_salary1$code2 == "27", c("Arts, Design, Entertainment, Sports, and Media"), state_salary1$major)
state_salary1$major <- ifelse(state_salary1$code2 == "29", c("Healthcare Practitioners"), state_salary1$major)
state_salary1$major <- ifelse(state_salary1$code2 == "31", c("Healthcare Support"), state_salary1$major)
state_salary1$major <- ifelse(state_salary1$code2 == "33", c("Protective Service"), state_salary1$major)
state_salary1$major <- ifelse(state_salary1$code2 == "35", c("Food Preparation and Serving"), state_salary1$major)
state_salary1$major <- ifelse(state_salary1$code2 == "37", c("Building and Grounds Cleaning and Maintenance"), state_salary1$major)
state_salary1$major <- ifelse(state_salary1$code2 == "39", c("Personal Care and Service"), state_salary1$major)
state_salary1$major <- ifelse(state_salary1$code2 == "41", c("Sales and Related"), state_salary1$major)
state_salary1$major <- ifelse(state_salary1$code2 == "43", c("Office and Administrative Support"), state_salary1$major)
state_salary1$major <- ifelse(state_salary1$code2 == "45", c("Farming, Fishing, and Forestry"), state_salary1$major)
state_salary1$major <- ifelse(state_salary1$code2 == "47", c("Construction and Extraction"), state_salary1$major)
state_salary1$major <- ifelse(state_salary1$code2 == "49", c("Installation, Maintenance, and Repair"), state_salary1$major)
state_salary1$major <- ifelse(state_salary1$code2 == "51", c("Production"), state_salary1$major)
state_salary1$major <- ifelse(state_salary1$code2 == "53", c("Transportation and Material Moving"), state_salary1$major)


```

```{r message=FALSE, warning=FALSE, echo=TRUE}

## Bar plot to show total employment count
state_salary2 <- state_salary1 %>% filter(Group=='total')

ggplot(state_salary2, aes(State, Emp.count)) + geom_bar(aes(fill=Emp.count), stat = "identity", width = 0.6, fill = "lightgreen") + geom_text(label = state_salary2$Emp.count) + coord_flip() + ggtitle("Total Employment count (Numbers)" ) + xlab(NULL) + ylab(NULL) + theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank())

```

Choropleth for US Annual Mean Salary, 2015, by STATE

```{r message=FALSE, warning=FALSE, echo=TRUE}
# removed 3 states as they are not mapped in choropleth
State_salary_choro <- state_salary2 %>% filter(State != "Guam" & State != "Puerto Rico" & State != "Virgin Islands") %>% select(State, A.mean.wage)

State_salary_choro$A.mean.wage <- as.numeric(State_salary_choro$A.mean.wage)
colnames(State_salary_choro) <- c("region", "value")
State_salary_choro$region <- str_to_lower(State_salary_choro$region)
state_choropleth(State_salary_choro, 
                 title  = "United States - Annual Mean Salary, 2015 by State", 
                 legend = "USD", num_colors = 1)

```


Load excel data sheet "US_Crime_2015" into R. This data sheet has US Crime data for various crime categories. The code converts the data into tall format using 'GATHER' command and plots a bar graph with States plotted on X axis and crime incident count for each crime category plotted on Y axis.


```{r message=FALSE, warning=FALSE}

## load and clean Crime data file
crime_file <- read_excel("US_Crime_2015.xls", skip=3, na="", col_names = TRUE, col_types = NULL)

## Add 'Area' value to each row

prevarea <- crime_file$Area    #Moved Area column data into a character vector


for(i in 2:nrow(crime_file))  # FOR loop to fill Area column value correctly whereever its blank or NA in the dataset
{ 
  if (is.na(prevarea[i])) {
    prevarea[i] <- prevarea[i-1]  # when area value blank, fill it from previous row area value
  } 
}

Crime_file1 <- data.frame(prevarea, crime_file)   

Crime_file1 <- Crime_file1 %>% filter(Crime_file1$Year == "2015.000000")  # select rows only for  2015


colnames(Crime_file1) <- c("State", "area", "year", "population", "violent.crime", "violent.crime.rate", "murder", "murder.rate", "rape", "rape.rate", "rape.legacy", "rape.legacy.rate", "robbery", "robbery.rate", "assault", "assault.rate", "property.crime", "property.crime.rate", "burglary", "burglary.rate", "larceny.theft", "larceny.theft.rate", "vehicle.theft", "vehicle.theft.rate","a","b")

Crime_file1$State <- tolower(Crime_file1$State) 

state_list <- data.frame(state_salary2$ST.code, state_salary2$State)  # Use state_salary2 dataframe to build state_list dataframe that will have 2 columns , state name and state code. 

colnames(state_list) <- c("state.code", "state")
state_list$state <- str_trim(tolower(state_list$state))  


Crime_file1$State <- str_trim(gsub(",","",str_trim(str_replace_all(Crime_file1$State,"[0-9]+",""))))    #clean Crime_file1 column to remove numeric digits from state name  


crime_file3 <- left_join(Crime_file1, state_list, by=c("State" = "state"))  # join crime table and state list table on state name

crime_file4 <- crime_file3 %>% filter(!is.na(state.code))    # select rows that have valid value for state code
crime_file4$State <- str_trim(crime_file4$State)
crime_file4$year <- fround(as.numeric(crime_file4$year), 0)   # remove decimals/zeros from column Year
crime_file4$population <- as.numeric(fround(as.numeric(crime_file4$population), 0)) # remove decimals/zeros from column population
crime_file4$violent.crime <- as.numeric(fround(as.numeric(crime_file4$violent.crime), 2))
crime_file4$violent.crime.rate <- as.numeric(fround(as.numeric(crime_file4$violent.crime.rate), 2))
crime_file4$murder <- as.numeric(fround(as.numeric(crime_file4$murder), 2))
crime_file4$murder.rate <- as.numeric(fround(as.numeric(crime_file4$murder.rate), 2))
crime_file4$rape <- as.numeric(fround(as.numeric(crime_file4$rape), 2))
crime_file4$rape.rate <- as.numeric(fround(as.numeric(crime_file4$rape.rate), 2))
crime_file4$rape.legacy <- as.numeric(fround(as.numeric(crime_file4$rape.legacy), 2))
crime_file4$rape.legacy.rate <- as.numeric(fround(as.numeric(crime_file4$rape.legacy.rate), 2))
crime_file4$robbery <- as.numeric(fround(as.numeric(crime_file4$robbery), 2))
crime_file4$robbery.rate <- as.numeric(fround(as.numeric(crime_file4$robbery.rate), 2))
crime_file4$assault <- as.numeric(fround(as.numeric(crime_file4$assault), 2))
crime_file4$assault.rate <- as.numeric(fround(as.numeric(crime_file4$assault.rate), 2))
crime_file4$property.crime <- as.numeric(fround(as.numeric(crime_file4$property.crime), 2))
crime_file4$property.crime.rate <- as.numeric(fround(as.numeric(crime_file4$property.crime.rate), 2))
crime_file4$burglary <- as.numeric(fround(as.numeric(crime_file4$burglary), 2))
crime_file4$burglary.rate <- as.numeric(fround(as.numeric(crime_file4$burglary.rate), 2))
crime_file4$larceny.theft <- as.numeric(fround(as.numeric(crime_file4$larceny.theft), 2))
crime_file4$larceny.theft.rate <- as.numeric(fround(as.numeric(crime_file4$larceny.theft.rate), 2))
crime_file4$vehicle.theft <- as.numeric(fround(as.numeric(crime_file4$vehicle.theft), 2))
crime_file4$vehicle.theft.rate <- as.numeric(fround(as.numeric(crime_file4$vehicle.theft.rate), 2))


### Crime data cleaning end ##########

crime_file5 <- crime_file4 %>% select(state.code, violent.crime, murder, rape, robbery, assault, property.crime, burglary, larceny.theft, vehicle.theft)

crime_file6 <- crime_file4 %>% select(state.code, violent.crime.rate, murder.rate, rape.rate, robbery.rate, assault.rate, property.crime.rate, burglary.rate, larceny.theft.rate, vehicle.theft.rate)

```

Converted Crime data frame into tall format using GATHER. This is needed to plot crime count by category.

```{r message=FALSE, warning=FALSE, echo=TRUE}

tall_crime_count <- gather(crime_file5, key="crime.category", 2:10, value = "count")

tall_crime_rate <- gather(crime_file6, key="crime.category", 2:10, value = "rate")


```

2015 CRIME INCIDENT COUNT PLOTTED ON Y-AXIS AND BAR IS PLOTTED FOR EACH STATE ON X-AXIS

```{r message=FALSE, warning=FALSE, echo=TRUE}

ggplot(tall_crime_count, aes(state.code, count)) +
  geom_bar(aes(fill = count), stat = "identity", colour="black", fill="skyblue") +
  ylab('Crime Incident Count') + xlab(NULL) +
  facet_wrap(~crime.category, scales='free') + theme(axis.text.x = element_text(angle=60, hjust=1))


```

2015 CRIME RATE (PER 100,000) PLOTTED ON Y-AXIS AND BAR IS PLOTTED FOR EACH STATE ON X-AXIS

```{r message=FALSE, warning=FALSE, echo=TRUE}

ggplot(tall_crime_rate, aes(state.code, rate)) +
  geom_bar(aes(fill = rate), stat = "identity", colour="black", fill="skyblue") +
  ylab('Crime Rate (per 100,000)') + xlab(NULL) +
  facet_wrap(~crime.category, scales='free') + theme(axis.text.x = element_text(angle=60, hjust=1))


```